Bayesian Nonparametric Neural Image Segmentation

### 2013-03-01 latex/submission.tex
updated author list, some more might come
added DDP and M-H sampler code

\title{Bayesian Nonparametric Neural Image Segmentation}
\author{Kui Tang and Frank Wood}
\date{\today}
\maketitle
\begin{abstract}
We develop a Bayesian nonparametric method to segment neurons in electron micrograph images of brain tissue. Our model imposes spatial locality on arbitrary features, automatically determines the number of segments in an image, and can be used in both unsupervised and semi-supervised settings. We introduce a novel geodesic distance feature which adaptively adjust segmentation to respect detected cell boundaries. Using data collected by Kasthuri et al., we show that our model can obtain plausible segmentations with no training data, while receiving a significant boost in its presence. Finally, our Bayesian nonparametric naturally extends to 3 dimensions, where neural columns of segments \emph{through} slices are modelled.
\end{abstract}

@@

\subsection{Geodesic Distance}
\label{sec:geodesic}
\begin{figure}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_small}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_dist}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_thresh}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_hist}
\caption{(1) Boundary Radon-like feature response; red marks the center from which the remaining pictures are measured. (2) Grayscale map of geodesic distance (of superpixels). (3) Threshold of geodesic distance $< 0.1$. Distribution of geodesic distances about the red center.\label{fig:geodesic}}
\end{figure}
Figures~\ref{fig:radon} and~\ref{fig:geodesic} illustrate that cells need not be elliptical, or even convex. All that can be said is that they are contained in a closed loop that is generally gives a distinct Radon-like boundary response. Therefore, shape constraints based on Euclidean distances are not justified. Instead, we want to encode the a priori belief that ``the cell body should not spill outside its membrane.'' Let us quantify this statement. Construct an undirected graph with pixels as vertices and pairs of adjacent pixels as edges ($(x,y)$ is an edge if $y$ is in the 1-pixel neighborhood of $x$). Each edge has a weight $w(i,j) = \left| b_i - b_j \right|$ where $b_k$ is the boundary Radon-like response at pixel $k$. We define the \emph{geodesic distance} $d(\+{x}_i, \+{x}_j)$ to be the total weight of the shortest path between pixels $i$ and $j$, which is computed with Johnson's algorithm. In practice, the $O(N^2)$ storage is prohibitive, so we instead we downsample the image into SLIC superpixels \cite{Achanta2012} and compute superpixel-based distances. Further, because individual cells are tiny relative to an entire slice, only local distances are needed, because far points must be outside the current cell. All paths crossing a cell boundary must traverse pixels with low, high, and again low boundary feature responses. Thus, the shortest path will also be one with a high weight. So if pixels $i$ and $j$ do not belong to the same cell, defined by the region enclosed by a loop of pixels with high boundary feature response, then $d(\+{x}_i, \+{x}_j)$ is high.

Recall that the cluster mean contains entries for the $xy$-coordinates $\+{\mu}_{k,1:2} = (\mu_{k,x}, \mu_{k,y})$. Fix a point $n$ and a cluster $k$. Then $d(\+{x}_n, \+{\mu}_{k,1:2})$ is exponentially distributed, favoring distances close to zero. Any point one the opposite side of a boundary from the cluster's center will be far away from zero, and hence have low probability. Geodesic distances to other points are ignored.

\section{Inference}
Inference is performed by Gibbs sampling, using Radford Neal's Algorithm 8 for non-conjugate priors \cite{Neal2000}. This is because the likelihood of $\+x$, which depends on the function $d$, is not a known conjugate-parametric form.\fxwarning{I'm not sure this is the right explanation.} We collapse $\+\pi$, but we do not collapse the other latent variables. Denote by $X_{-n}$ the set of all $\+x_n$ with point $n$ removed, by $X_k = \left\{ \+x_n \in X : z_n = k \right\}$ and $N_k = | X_k |$. Then we derive the conditionals by finding the Markov blankets of each node and using Bayes' rule. But because the likelihood of $\+x_n$ is not a standard parametric form, conditionals involving $X_k$ and $\+\mu_k$ cannot be sampled explicitly; we can only compute their likelihoods up to proportionality and resort to Metropolis-Hastings sampling.
\begin{align}\label{eq:conds}
p(z_n = k | \+\pi, \+x_n, \+\mu_k, \Lambda_k) &\propto p(z_n = k | \+\pi) \mathcal{N}(\+x_n | \+\mu_k, \Lambda_k) \label{eq:z}\\
p(\gamma_k | X_k, \+\mu_k, \beta_0, \gamma_0) &= \text{Gamma}\left( \gamma_k | \beta_0 + N_k, \gamma_0 + \sum_{\+x \in X_k} d(\+x, \+\mu_k) \right) \label{eq:gamma}\\
p(\Lambda_k | X_k, \+\mu_k, \gamma_k, \Lambda_0, \nu) &\propto  p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \text{Wishart}(\Lambda_k | \Lambda_0, \nu) \label{eq:lambda}\\
p(\+\mu_k | X_k, \Lambda_k, \gamma_k, \+\mu_0, \kappa) &\propto p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \mathcal{N}(\+\mu_k | \+\mu_0, (\kappa\Lambda)^{-1}) \label{eq:mu}
\end{align}
Note that $\+\pi$ remains collapsed; to evaluate \eqref{eq:z}, see \eqref{eq:mdp}.

@@

\For{$n := 1:N$}
  \If{$N_{z_n} = 1$}
    \State $\left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right) := \left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right)$ 
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+2, \ldots, K+m\}$
  \Else
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+1, \ldots, K+m\}$
  \EndIf
  \State Draw $z_n$ from 
  \[
  p(z_{n}=k|\+{z}_{-n},X_{-n},\alpha,\left\{ \+\mu_k,\Lambda_k,\gamma_k\ \right\}_{k=1}^{\infty}) \propto \begin{cases}
  \frac{N_{k,-n}}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & k < K\\
  \frac{\alpha / m}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & \text{otherwise}
  \end{cases} \label{eq:nealalg8}
  \]
  \State Delete parameters for classes $k$ with $N_k = 0$ and relabel the $z_n$ to remove gaps.
\EndFor

>>>>

\title{Bayesian Nonparametric Neural Image Segmentation}
\author{Kui Tang, $\sigma$(Joshua T.~Vogelstein, Jeff Lichtman, Bobby Kasthuri, Randal Burns, 
\\ Priya Manavalan) Frank Wood}
\date{\today}
\maketitle
\begin{abstract}
We develop a Bayesian nonparametric method to segment neurons in electron micrograph images of brain tissue. Our model imposes spatial locality on arbitrary features, automatically determines the number of segments in an image, and can be used in both unsupervised and semi-supervised settings. We introduce a novel geodesic distance feature which adaptively adjust segmentation to respect detected cell boundaries. Using data collected by Kasthuri et al., we show that our model can obtain plausible segmentations with no training data, while receiving a significant boost in its presence. Finally, our Bayesian nonparametric naturally extends to 3 dimensions, where neural columns of segments \emph{through} slices are modelled.
\end{abstract}

@@

\subsection{Geodesic Distance}
\label{sec:geodesic}
\begin{figure}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_small}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_dist}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_thresh}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_hist}
\caption{(1) Boundary Radon-like feature response; red marks the center from which the remaining pictures are measured. (2) Grayscale map of geodesic distance (of superpixels). (3) Threshold of geodesic distance $< 0.1$. Distribution of geodesic distances about the red center.\label{fig:geodesic}}
\end{figure}
Figures~\ref{fig:radon} and~\ref{fig:geodesic} illustrate that cells need not be elliptical, or even convex. All we can say is that they are contained in a closed loop that is generally gives a distinct Radon-like boundary response. Therefore, shape constraints based on Euclidean distances are not justified. Instead, we want to encode the a priori belief that ``the cell body should not spill outside its membrane.'' Let us quantify this statement. Construct an undirected graph with pixels as vertices and pairs of adjacent pixels as edges ($(x,y)$ is an edge if $y$ is in the 1-pixel neighborhood of $x$). Each edge has a weight $w(i,j) = \left| b_i - b_j \right|$ where $b_k$ is the boundary Radon-like response at pixel $k$. We define the \emph{geodesic distance} $d(\+{x}_i, \+{x}_j)$ to be the total weight of the shortest path between pixels $i$ and $j$, which is computed with Johnson's algorithm. In practice, the $O(N^2)$ storage is prohibitive, so we instead we downsample the image into SLIC superpixels \cite{Achanta2012} and compute superpixel-based distances. Further, because individual cells are tiny relative to an entire slice, only local distances are needed, because far points must be outside the current cell. All paths crossing a cell boundary must traverse pixels with low, high, and again low boundary feature responses. Thus, the shortest path will also be one with a high weight. So if pixels $i$ and $j$ do not belong to the same cell, defined by the region enclosed by a loop of pixels with high boundary feature response, then $d(\+{x}_i, \+{x}_j)$ is high.

Recall that the cluster mean contains entries for the $xy$-coordinates $\+{\mu}_{k,1:2} = (\mu_{k,x}, \mu_{k,y})$. Fix a point $n$ and a cluster $k$. Then $d(\+{x}_{n,1:2}, \+{\mu}_{k,1:2})$ is exponentially distributed, favoring distances close to zero. Any point one the opposite side of a boundary from the cluster's center will be far away from zero, and hence have low probability. Geodesic distances to other points are ignored.

\section{Inference}
Inference is performed by Gibbs sampling, using Radford Neal's Algorithm 8 for non-conjugate priors \cite{Neal2000}. This is because the likelihood of $\+x$, which depends on the function $d$, is not a known conjugate-parametric form.\fxwarning{I'm not sure this is the right explanation.} To use this sampling method, we collapse $\+\pi$, but we do not collapse the other latent variables. Denote by $X_{-n}$ the set of all $\+x_n$ with point $n$ removed, by $X_k = \left\{ \+x_n \in X : z_n = k \right\}$ and $N_k = | X_k |$. Then we derive the conditionals by finding the Markov blankets of each node and using Bayes' rule:
\begin{align}\label{eq:conds}
p(z_n = k | \+\pi, \+x_n, \+\mu_k, \Lambda_k, \gamma_k) &\propto p(z_n = k | \+\pi) \mathcal{N}(\+x_n | \+\mu_k, \Lambda_k) \text{Exponential}\left( d(\+x_{1:2}, \+\mu_{k,1:2}) | \gamma_k \right) \label{eq:z}\\
p(\gamma_k | X_k, \+\mu_k, \beta_0, \gamma_0) &= \text{Gamma}\left( \gamma_k | \beta_0 + N_k, \gamma_0 + \sum_{\+x \in X_k} d(\+x, \+\mu_k) \right) \label{eq:gamma}\\
p(\Lambda_k | X_k, \+\mu_k, \gamma_k, \Lambda_0, \nu) &\propto  p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \text{Wishart}(\Lambda_k | \Lambda_0, \nu) \label{eq:lambda}\\
p(\+\mu_k | X_k, \Lambda_k, \gamma_k, \+\mu_0, \kappa) &\propto p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \mathcal{N}(\+\mu_k | \+\mu_0, (\kappa\Lambda)^{-1}) \label{eq:mu}
\end{align}

Because the likelihood of $\+x_n$ is not a standard parametric form, conditionals involving $X_k$ and $\+\mu_k$ cannot be sampled explicitly; we can only compute their likelihoods up to proportionality and resort to Metropolis-Hastings sampling. Further, $\+\pi$ remains collapsed; to evaluate \eqref{eq:z}, see \eqref{eq:mdp}.

@@

\For{$n := 1:N$}
  \If{$N_{z_n} = 1$}
    \State $\left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right) := \left( \+\mu_{z_n}, \Lambda_{z_n}, \gamma_{z_n} \right)$ 
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+2, \ldots, K+m\}$
  \Else
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+1, \ldots, K+m\}$
  \EndIf
  \State Draw $z_n$ from 
  \[
  p(z_{n}=k|\+{z}_{-n},X_{-n},\alpha,\left\{ \+\mu_k,\Lambda_k,\gamma_k\ \right\}_{k=1}^{\infty}) \propto \begin{cases}
  \frac{N_{k,-n}}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & k < K\\
  \frac{\alpha / m}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & \text{otherwise}
  \end{cases} \label{eq:nealalg8}
  \]
  \State Delete parameters for classes $k$ with $N_k = 0$ and relabel the $z_n$ to remove gaps.
\EndFor
